Binning Strategies and Related Loss for Large Data

### 2015-03-29 Paper.Rnw
Complete intro to section 5 with scatterplot of simulated data

\ktm{ Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices. Bivariate datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.  These distributions were selected for their variety in terms of shape, center and spread. 100,000 observation pairs were simulated for each set of simulated data, and the x and y values were drawn independently from univariate distributions; Exponental(11), Normal(50,$11^2$) and Uniform(0,100), respectively.  The parameters were selected to have data values roughly span the region $[0,100]^2$.  }

\ktm{ These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places. Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles.  To observe binning loss under more realistic conditions, three corresponding data sets were also created by rounding the values from the original simulated data sets to the nearest even number. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded. }

\ktm{
\begin{itemize}
\item Binning and frequency binning involve choices (standard/random/quantile, origin, binwidth, number of colors) want to make optimal choices given goals of binning.
\item 3 simulated data sets: from bivariate Exponential, Normal, Uniform
\item Since most data recorded to rough resolution we also assess loss when each dataset is rounded to nearest even value.  
\end{itemize}
}

>>>>

\ktm{ Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices. Bivariate datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.  These distributions were selected for their variety in terms of shape, center and spread. 100,000 observation pairs were simulated for each set of simulated data, and the x and y values were drawn independently from univariate distributions; Exponental(11), Normal(50,$11^2$) and Uniform(0,100), respectively.  The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}. }

<<SimDataScatterplots,echo=F,include=T,eval=T,fig.width=8, fig.height=6, out.width='.8\\linewidth', fig.pos='h',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Scatterplots of fine and coarse versions of the simulated bivariate data.">>= 
### Try 2d binning with MVN data
n=100000
set.seed(31415)
unifdat <- data.frame(x= runif(n, min=0, max=100),
                      y= runif(n, min=0, max=100),
                      type="Uniform",
                      record = "Fine")
normdat <- data.frame(x= rnorm(n, mean=50, sd=11),
                     y= rnorm(n, mean=50, sd=11),
                     type="Normal",
                     record = "Fine")
expdat <- data.frame(x= rexp(n, .11),
                     y= rexp(n, .11),
                     type="Exponential",
                     record = "Fine")
datlist <- list(unifdat,normdat,expdat)
### round all data to resolution alpha
alpha <- 2
for(i in 1:3){
  datlist[[i]]$roundx <- ((datlist[[i]]$x + 1)%/%alpha)*alpha
  datlist[[i]]$roundy <- ((datlist[[i]]$y + 1)%/%alpha)*alpha
}

# plot the original data
# make rounded data find overlapping points to make plot rendering faster
dat <- rbind(datlist[[1]],datlist[[2]],datlist[[3]])
smallround <- ddply(dat,.(type,record,roundx,roundy),summarise,
      overlap = length(roundx))
smallround <-cbind(data.frame(x=smallround$roundx,y=smallround$roundy), 
                   smallround[,1:4])
smallround$record <- "Coarse"
alldat <- rbind(dat,smallround)

qplot(x,y,geom="point",data=alldat,facets=record~type, size=I(1)) + 
  theme_bw() + 
  #ggtitle("Scatterplots of Simulated Data") +
  xlab("X") + ylab("Y") + 
  scale_x_continuous(breaks=c(0,25,50,75,100)) + 
  scale_y_continuous(breaks=c(0,25,50,75,100))
@

\ktm{ These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places. Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles. To observe binning loss under more realistic conditions, three corresponding data sets were also created by rounding the values from the original simulated data sets to the nearest even number. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded. }
